Differential Expression - HBCC - DLPFC - Dx - Sex

### 2020-08-18 individual-cohorts/HBCC_DLPFC_DE.Rmd
dx*sex interaction, revised model and contrasts, fix conflict

# Test Dx-Sex
COVARIATES <- COVARIATES %>%
  tibble::rownames_to_column(var = "samples") %>% 
  tidyr::unite("Dx_Sex",
               dplyr::all_of(c("Dx", "Reported_Gender")),
               remove = TRUE) %>% 
  dplyr::mutate(`Dx_Sex` = as.factor(`Dx_Sex`)) %>% 
  tibble::column_to_rownames(var = "samples")

# Set covariates to adjust for from model selection
adjust.covars <- c("Dx_Sex", "ageOfDeath", "RIN", "IntronicRate", "IntragenicRate", "IntergenicRate")
```

### Normalisation - Dx_Sex
1. Dx_Sex is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis)

```{r iterative.normalisation, results='asis'}
writeLines(glue::glue('Using following covariates in the model:',
                 glue::glue_collapse(adjust.covars, sep = ", ", last = " and "),
                 ' as fixed effects.'))

@@

# Post adjusted formula
formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + "))

@@

formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(formula_for_variance_fractions, sep = " + "))

@@

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION)
```

### Residual calculation
Calculate weighted residuals and add back "Dx_Sex" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx_Sex", colnames(ADJUSTED.FIT$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
```

@@

```{r decompose.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx_Sex'), drop = FALSE])

@@

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx_Sex'), drop = FALSE]))
```

@@

### Differential expression analysis (SCZ - Control - Other)
Genes that are differentially expressed at an FDR <= 0.05 are:
```{r set-formula}
formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + "))    
```

@@

```{r diffExp, fig.height=10, fig.width=15}
L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx_SexControl_Female", "Dx_SexControl_Male"))
L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx_SexSCZ_Female", "Dx_SexSCZ_Male"))
L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx_SexOther_Female", "Dx_SexOther_Male"))

# Manually set comparison between SCZ+Other vs. Control 
L = cbind(L1, L2, L3, 
          data.frame(L4 = c(1,0,-0.5, 0, -0.5, 0, 0, 0, 0, 0, 0)),
          data.frame(L5 = c(0, 1, 0, -0.5, 0, -0.5, 0, 0, 0, 0, 0)))

contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
              glue::glue_collapse(names(L2[L2 != 0]), "-"),
              glue::glue_collapse(names(L3[L3 != 0]), "-"),
              glue::glue("Dx_SexControl_Female-Dx_SexOther_FemaleDx_SexSCZ_Female"),
              glue::glue("Dx_SexControl_Male-Dx_SexOther_MaleDx_SexSCZ_Male")
)

@@

# Get gene lengths
counts = 'syn21886235'
ALL_USED_IDs = counts
counts = downloadFile(counts) %>% data.frame()
gene_lengths = counts[,c("Geneid", "Length")] %>% 
  dplyr::rename(gene_id = Geneid)
rm(counts)

@@

# Get differential expression
DE = lapply(glue::glue("L{1:5}"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>% 
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) <- contrast

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::unite(Comparison, `ref.state`, `to.state`, sep = '_vs_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths)
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

>>>>

# Set covariates to adjust for from model selection
interaction <- c("Dx", "Reported_Gender")
adjust.covars <- c("ageOfDeath", "RIN", "IntronicRate", "IntragenicRate", "IntergenicRate")
```

### Normalisation - Dx*Sex
1. Dx*Sex is chosen as the primary variable of interest

```{r iterative.normalisation, results='asis'}
writeLines(glue::glue('Using following covariates in the model:',
                      glue::glue_collapse(interaction, sep = ", "),
                      ', ',
                      glue::glue_collapse(adjust.covars, sep = ", ", last = " and "),
                      ' as fixed effects.'))

@@

# Post adjusted formula
formula <- glue::glue("~ ", glue::glue_collapse(interaction, sep = "*")," + ", glue::glue_collapse(adjust.covars, sep = " + "))

@@

# Manually specify interaction
interaction <- c(interaction, "Dx:Reported_Gender")
interaction_variance <- glue::glue("(1|{interaction})")
all <- c(interaction_variance, formula_for_variance_fractions)

formula_for_variance_fractions <- glue::glue("~ ", glue::glue_collapse(all, sep = " + "))

@@

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals(ADJUSTED.FIT, CQN.GENE_EXPRESSION)
```
`r formula`
### Residual calculation
Calculate weighted residuals and add back "Dx:Sex" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("DxOther:Reported_GenderMale", "DxSCZ:Reported_GenderMale", colnames(ADJUSTED.FIT$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
```

@@

```{r decompose.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx', 'Reported_Gender'), drop = FALSE])

@@

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx','Reported_Gender'), drop = FALSE]))
```

@@

### Differential expression analysis (SCZ - Control - Other)
Genes that are differentially expressed at an FDR <= 0.05 are:
```{r set-formula}
formula <- glue::glue("~ 0 + Dx_Sex + ", glue::glue_collapse(adjust.covars, sep = " + "))
```

@@

Dx_Sex is 6 levels, SCZ_Female, SCZ_Male, Other_Female, Other_Male, Control_Female, Control_Male

```{r diffExp, fig.height=10, fig.width=15}
# Test Dx-Sex
COVARIATES <- COVARIATES %>%
  tibble::rownames_to_column(var = "samples") %>% 
  tidyr::unite("Dx_Sex",
               dplyr::all_of(c("Dx", "Reported_Gender")),
               remove = TRUE) %>% 
  dplyr::mutate(`Dx_Sex` = as.factor(`Dx_Sex`)) %>% 
  tibble::column_to_rownames(var = "samples")

L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx_SexControl_Male", "Dx_SexControl_Female"))
L2 = c(1, -1, 0, 0, -1, 1, 0, 0, 0, 0, 0)
L3 = c(1, -1, -1, 1, 0, 0, 0, 0, 0, 0, 0)

# Manually set comparison between SCZ+Other vs. Control 
L = cbind(L1, L2, L3)

contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
              glue::glue("Dx_SexSCZ_MaleDx_SexControl_Male-Dx_SexSCZ_FemaleDx_SexControl_Female"),
              glue::glue("Dx_SexOther_MaleDx_SexControl_Male-Dx_SexOther_FemaleDx_SexControl_Female")
)

@@

# Get gene lengths
counts = 'syn21886235'
ALL_USED_IDs = c(counts, ALL_USED_IDs)
counts = downloadFile(counts) %>% data.frame()
gene_lengths = counts[,c("Geneid", "Length")] %>% 
  dplyr::rename(gene_id = Geneid)
rm(counts)

@@

# Get differential expression
DE = lapply(glue::glue("L{1:3}"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>% 
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) <- contrast

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx_Sex','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::unite(Comparison, `ref.state`, `to.state`, sep = '_vs_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths)
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

### Source Code
`r thisFile`
